---
title: "Non-Parametric Tests in R"
author: "Odia"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
alpha <- 0.05
```

# Non-Parametric Tests

Non-parametric tests are useful when assumptions of parametric tests are not satisfied, especially with non-normal data, ordinal data, outliers, or small samples.

Our pattern remains: business scenario, EDA, descriptive analysis, hypotheses, assumptions, test, decision, interpretation, recommendation, and lessons learned.

## Module 1: Mann-Whitney U Test

### Business Scenario

FreshMart wants to know whether purchase amounts differ between two independent campaigns when normality is not supported.

### Load and explore

```{r}
marketing_data <- read.csv("mann_whitney_data.csv")

head(marketing_data)
str(marketing_data)
summary(marketing_data)
table(marketing_data$Campaign)
```

### Groups

```{r}
old <- subset(
  marketing_data,
  Campaign == "Old"
)$Purchase_Amount

new <- subset(
  marketing_data,
  Campaign == "New"
)$Purchase_Amount
```

### Descriptive statistics

```{r}
mean(old)
median(old)
mean(new)
median(new)
sd(old)
sd(new)
```

### Normality

```{r}
shapiro.test(old)
shapiro.test(new)
```

### Hypotheses

H0: The two groups have the same distribution.

H1: The two groups have different distributions.

### Test

```{r}
mw_result <- wilcox.test(
  new,
  old,
  alternative = "two.sided",
  exact = FALSE,
  conf.int = TRUE
)

mw_result
```

### Decision

```{r}
if (mw_result$p.value < alpha) {
  print("Reject H0")
} else {
  print("Fail to Reject H0")
}
```

### Interpretation

A significant result indicates evidence that the two independent groups differ in distribution.

### Business Recommendation

Use the result to compare campaigns when the independent t-test assumptions are not suitable.

### What I Learned

- Mann-Whitney is a non-parametric alternative to the independent t-test.
- It uses ranks.
- The observations must be independent.

---

## Module 2: Wilcoxon Signed-Rank Test

### Business Scenario

FreshMart measures the same employees before and after training, but the differences are not normally distributed.

### Load and explore

```{r}
paired_data <- read.csv("wilcoxon_signed_rank_data.csv")

head(paired_data)
str(paired_data)
summary(paired_data)
```

### Differences

```{r}
paired_data$Difference <-
  paired_data$After -
  paired_data$Before

mean(paired_data$Difference)
median(paired_data$Difference)
sd(paired_data$Difference)

shapiro.test(paired_data$Difference)
```

### Hypotheses

H0: The median paired difference is 0.

H1: The median paired difference is not 0.

### Test

```{r}
wilcox_result <- wilcox.test(
  paired_data$After,
  paired_data$Before,
  paired = TRUE,
  alternative = "two.sided",
  exact = FALSE,
  conf.int = TRUE
)

wilcox_result
```

### Decision

```{r}
if (wilcox_result$p.value < alpha) {
  print("Reject H0")
} else {
  print("Fail to Reject H0")
}
```

### Interpretation

A significant result indicates evidence of a systematic change between paired measurements.

### Business Recommendation

Use the result to assess an intervention when normality of the paired differences is not supported.

### What I Learned

- Wilcoxon Signed-Rank is the non-parametric counterpart to the paired t-test.
- It works with paired observations and signed ranks.
- Zero represents no paired change.

---

## Module 3: Kruskal-Wallis Test

### Business Scenario

FreshMart wants to compare purchase amounts across three or more cities, but the data are not suitable for one-way ANOVA.

### Load and explore

```{r}
marketing_data <- read.csv("kruskal_wallis_data.csv")

head(marketing_data)
str(marketing_data)
summary(marketing_data)
table(marketing_data$City)
```

### Hypotheses

H0: The groups come from the same distribution.

H1: At least one group differs.

### Test

```{r}
kruskal_result <- kruskal.test(
  Purchase_Amount ~ City,
  data = marketing_data
)

kruskal_result
```

### Decision

```{r}
if (kruskal_result$p.value < alpha) {
  print("Reject H0")
} else {
  print("Fail to Reject H0")
}
```

### Post hoc analysis

If significant:

```{r}
pairwise.wilcox.test(
  marketing_data$Purchase_Amount,
  marketing_data$City,
  p.adjust.method = "bonferroni"
)
```

### Interpretation

A significant Kruskal-Wallis result indicates that at least one group differs in distribution.

### Business Recommendation

Investigate the specific groups identified by post hoc testing before implementing location-wide decisions.

### What I Learned

- Kruskal-Wallis is the non-parametric counterpart to one-way ANOVA.
- It compares three or more independent groups.
- It uses ranks.
- A significant result requires post hoc comparisons.

---

## Module 4: Chi-Square Test of Independence

### Business Scenario

FreshMart wants to know whether membership status is associated with campaign response.

### Load and explore

```{r}
marketing_data <- read.csv("chi_square_data.csv")

head(marketing_data)
str(marketing_data)
summary(marketing_data)

table(
  marketing_data$Membership,
  marketing_data$Response
)
```

### Observed frequencies

```{r}
observed <- table(
  marketing_data$Membership,
  marketing_data$Response
)

observed
```

### Hypotheses

H0: Membership status and campaign response are independent.

H1: Membership status and campaign response are associated.

### Test

```{r}
chi_result <- chisq.test(observed)
chi_result
```

### Expected frequencies

```{r}
chi_result$expected
```

Expected frequencies should generally be sufficiently large for the chi-square approximation to be reliable. For small tables with small expected counts, consider Fisher's exact test.

### Decision

```{r}
if (chi_result$p.value < alpha) {
  print("Reject H0")
} else {
  print("Fail to Reject H0")
}
```

### Interpretation

A significant result indicates evidence of an association between the categorical variables. It does not prove causation.

### Business Recommendation

If an association exists, investigate which membership groups respond differently and use the result to improve campaign targeting.

### What I Learned

- Chi-square tests categorical variables.
- It compares observed and expected frequencies.
- A significant result indicates association, not causation.
